---
title: "Blocked Practice Stimuli Analysis"
output:
  html_document: default
  word_document: default
  pdf_document: default
  self_contained: true
date: "2026-04-14"
---

<style>
pre {
  margin-top: 50px;
  margin-bottom: 50px;
}

body {
  font-size: 20px;
}
</style>

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(purrr)
library(dplyr)
library(data.table)
library(readxl)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(ggthemes)
library(patchwork)
library(tibble)
library(corrplot)


```

## Import & Organize Stimuli Sheet
```{r}
# Import
stimuli <- suppressMessages(
  read_excel("../stimuli/660CorpusMasterSheet.xlsx"))

# Grab relevant columns, sort by Category, mark and exclude items w/ no category label 
stimuli_clean<-stimuli %>%select(Name, Category) %>% arrange(Category)%>%filter(Category!="NA")

# Split Categories into Separate Files 
#stimuli_split <- split(stimuli_clean, stimuli_clean$Category)

# Create main output directory
#dir.create("stimuli_by_category", showWarnings = FALSE)

# Write separate csv files, create subfolder for each category
#for (cat in names(stimuli_split)) {
  #dir.create(file.path("stimuli_by_category", cat), showWarnings = FALSE)
  #df_out <- stimuli_split[[cat]][ , !(names(stimuli_split[[cat]]) %in% "Category") ]   # remove Category column for data processing
 # write.csv(
  #  df_out,
  #  file = file.path("stimuli_by_category", cat, "vocab.csv"),
  #  row.names = FALSE)}

```


First, I imported the stimuli excel sheet and split the stimuli into vocabulary lists for each category. I then extracted 512-dimensional word embeddings from *Universal Sentence Encoder* for each word in each category, including the category label itself. This is a type of distributional semantic model, which represents text as numerical vectors such that semantically similar inputs are located near one another in a high-dimensional space. These embeddings look like this (with 512 dimensions (rows) per embedding):

```{r}
mammals_embeddings<-read.csv("stimuli_by_category/Mammals/USE_embeddings.csv")
mammals_embeddings[1:6,1:6]
```

The resulting lists/embeddings were then run through get_similarities.py to create semantic matrices for each category, including N x N matrices for calculating interitem similarities between all words in the list and 1 x N matrices to get similarity between all items and the category label itself. Similarity between each pair is calculated as the cosine of the angle between the two numerical vectors representing each word. For example:

```{r}
# Example
cat_vec <- c(1.0, 0.0)
dog_vec <- c(0.63, sqrt(1 - 0.63^2))  # unit vector with cos(theta)=0.63

# Cosine similarity function
cos_sim <- function(a, b) {
  sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)))}

#Plot vectors
sim <- cos_sim(cat_vec, dog_vec)
angle_rad <- acos(sim)
angle_deg <- angle_rad * 180 / pi

vec_df <- data.frame(
  x = c(0, 0),
  y = c(0, 0),
  xend = c(cat_vec[1], dog_vec[1]),
  yend = c(cat_vec[2], dog_vec[2]),
  word = c("cat", "dog"))

# Arc to show the angle
theta <- seq(0, angle_rad, length.out = 100)
arc_df <- data.frame(
  x = 0.25 * cos(theta),
  y = 0.25 * sin(theta))

#plot
ggplot() +
  geom_segment(data = vec_df,
               aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.25, "cm")),
               linewidth = 1.2) +
  geom_text(data = vec_df,
            aes(x = xend, y = yend, label = word),
            vjust = -0.7,
            size = 5) +
  geom_path(data = arc_df, aes(x = x, y = y), linewidth = 1) +
  annotate("text",
           x = 0.43, y = 0.12,
           label = paste0("θ = ", round(angle_deg, 1), "°"),
           size = 4.5) +
  annotate("text",
           x = 0.55, y = -0.15,
           label = paste0("cos(θ) = ", round(sim, 2)),
           size = 5) +
  coord_fixed(xlim = c(-0.1, 1), ylim = c(-0.2, 1.1)) +
  theme_classic()+annotate("text",
         x = 0.5, y = 0.95,
         label = "cos(theta) == (cat %.% dog) / (sqrt(cat %.% cat) * sqrt(dog %.% dog))",
         parse = TRUE,
         size = 3)


```

 For example, the N x N item similarity matrix for Mammals looks like this (only a subset for illustrative purposes):

```{r}
mammals_ex_items<-read.csv("stimuli_by_category/Mammals/USE_semantic_matrix_items.csv")%>%rename(" "="X")
mammals_ex_items[1:6,1:6]
```

While the 1 x N category label-item similarity matrix looks like this:

```{r}
mammals_ex_label<-read.csv("stimuli_by_category/Mammals/USE_semantic_matrix_category_label.csv")%>%rename(" "="X")
knitr::kable(mammals_ex_label)
```


These matrices were then run through a function to get average similarities for each matrix type (skipping diagonals for average interitem similarity).

## Looking At Average Similarities

We now have three types of similarity. 1) Category Norm/Fit, from original ratings, 2) Similarity between category label and each item, derived from word embeddings, and 3) Pairwise similarities between items in each category, derived from word embeddings. First, we will calculate the average of each type of similarity for each category and combine into one data frame. 

```{r}
#calculate average similarity (ie, how well given item fits in category) from original rating data
stimuli_avg <- stimuli %>%
  select(Name, Category, "Category Norm") %>%
  rename(CategoryNorm = "Category Norm") %>%
  filter(Category != "NA") %>%
  arrange(Category) %>%
  mutate(
    CategoryNorm_scaled = (CategoryNorm - min(CategoryNorm, na.rm = TRUE)) / #min/max scaled fit score for later comparison
      (max(CategoryNorm, na.rm = TRUE) - min(CategoryNorm, na.rm = TRUE))) %>%
  group_by(Category) %>%
  summarize(
    Average_Similarity = mean(CategoryNorm_scaled, na.rm = TRUE),       # scaled mean
    SD_Similarity = sd(CategoryNorm_scaled, na.rm = TRUE),              # scaled SD
    Average_Similarity_raw = mean(CategoryNorm, na.rm = TRUE),          # raw mean
    SD_Similarity_raw = sd(CategoryNorm, na.rm = TRUE)                  # raw SD
  ) %>%
  mutate(Similarity_Type = "Rating")

#import DSM derived item x label similarities for each category & combine into one df
base_dir <- "stimuli_by_category"

item_similarities_long <- list.dirs(base_dir, recursive = FALSE) %>%
  map_dfr(function(cat_path) {
    cat_name <- basename(cat_path)
    file_path <- file.path(cat_path, "USE_semantic_matrix_category_label.csv")
    if (!file.exists(file_path)) {
      message("Skipping ", cat_name)
      return(NULL)}
    # read matrix (row names = label word)
    df <- read.csv(file_path, row.names = 1, check.names = FALSE)
    df %>%
      tibble::rownames_to_column("Label_Word") %>%
      pivot_longer(
        cols = -Label_Word,
        names_to = "Item",
        values_to = "Similarity"
      ) %>%
      mutate(Category = cat_name)
  }) %>%
  select(Category, Item, Similarity)

#import AVERAGED label x item and interitem similarities 
average_similarities<-read.csv("category_average_similarity_combined.csv")%>%rename("Similarity_Type"="Matrix_Type")


#link data
average_similarities_comp<-average_similarities%>%rbind(stimuli_avg%>%select(-Average_Similarity_raw,-SD_Similarity_raw,-SD_Similarity))%>%
  pivot_wider(
    names_from = Similarity_Type,
    values_from = Average_Similarity)
```


Now, let's look at average category fit from the original ratings. Note that 1 = Fits very well into category, 4 = Very poor fit. We can then compare them to the average label-item and item-item similarities.

```{r}
#average category fit ratings
avg_ratings<-stimuli_avg %>% ggplot(aes(x = Category, y = Average_Similarity_raw,fill=Category)) +
  geom_col() +
  geom_errorbar(aes(ymin = Average_Similarity_raw - SD_Similarity_raw,
                    ymax = Average_Similarity_raw + SD_Similarity_raw),
                width = 0.2) + scale_y_reverse()+
  ylim(4,0) + ylab("Avg Rating (Poorest Fit -> Best Fit)")+ ggtitle("Avg Category Norm Ratings")+
  theme_classic() +theme(legend.position="none")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#average label x item similarities (DSM derived)
avg_label<-average_similarities_comp %>% mutate(SD_Similarity = sd(Label, na.rm = TRUE))%>% ggplot(aes(x = Category, y = Label,fill=Category)) +
  geom_col() +
  geom_errorbar(aes(ymin = Label - SD_Similarity,
                    ymax = Label + SD_Similarity),
                width = 0.2) +
   ylab("Avg Similarity (Poorest Fit -> Best Fit)") + ggtitle("Avg Category Label x Item Similarity")+
  theme_classic() +theme(legend.position="none")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#average inter-item similarity across categories
avg_interitem<-average_similarities_comp %>% mutate(SD_Similarity = sd(Interitem, na.rm = TRUE))%>% ggplot(aes(x = Category, y = Interitem,fill=Category)) +
  geom_col() +
  geom_errorbar(aes(ymin = Interitem - SD_Similarity,
                    ymax = Interitem + SD_Similarity),
                width = 0.2) +
  ylab("Avg Similarity (Poorest Fit -> Best Fit)") +ggtitle("Avg Item x Item Similarity")+
  theme_classic() +theme(legend.position="none")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


#compare distributions
avg_ratings
avg_label
avg_interitem
```
<div style="clear: both;"></div>
## Correlations 

We can then look at the correlation between the average Category Norm fit ratings and the embedding derived similarities. Note that min/max scaling was applied to the ratings to convert them to a 0-1 scale and plot axis was reversed to appropriately map onto embedding-derived similarities.

A note about interpretation. In the context of distributional semantic models, similarity follows Firth's quote *“You shall know a word by the company it keeps.”* So words are similar to the extent that they appear in similar contexts and are used in a similar manner within the text corpora the models are trained on. So when thinking about the similarity between the category label and each word, its important to consider that superordinate category labels are not used in the same manner as exemplar labels. So while the similarity between the label and exemplar is informative, it may be more informative to consider the average inter-item similarities as a measure of **cohesion** within a given category, which may provide more fine-grained information than the category fit ratings. It is arguably this type of cohesion that is driving stronger interference effects, such that it could be used for stimulus selection.

```{r}
#Correlation between average (scaled) ratings of category fit vs average DSM-derived similarity between category label and items
label_correlation <- average_similarities_comp %>%
  ggplot(aes(x = Label, y = Rating)) +
  geom_point() +
  scale_y_reverse() +
  geom_text_repel(aes(label = Category)) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Avg Category Label x Item Similarity") +
  ylab("Avg Category Fit Rating (Scaled)") + ggtitle("Correlation Between Avg Category Norm Ratings & Avg Label x Item Similarities")+
  theme_classic()

label_correlation

#Correlation between average (scaled) ratings of item x category fit vs average DSM-derived interitem similarity (cohesion)
interitem_corr <- average_similarities_comp %>%
  ggplot(aes(x = Interitem, y = Rating)) +
  geom_point() +
  scale_y_reverse() +
  geom_text_repel(aes(label = Category)) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Avg Interitem Similarity") +
  ylab("Avg Category Fit Rating (Scaled)") + ggtitle("Correlation Between Avg Category Norm Ratings & Avg Item x Item Similarities")+
  theme_classic()

interitem_corr

#may be best to choose items based on 1) avg fit between item and category label and 2) Category cohesion (avg interitem similarity)
```

## Distributions
Lastly, we can also look at outliers in each category to mark weak items. Note this is only for the ratings-derived and label x item similarities.

```{r}
#Item-wise distribution of normative ratings across categories
#ID outliers
stimuli_outliers <- stimuli %>%
  filter(Category != "NA") %>%
  group_by(Category) %>%
  mutate(
    Q1 = quantile(`Category Norm`, 0.25, na.rm = TRUE),
    Q3 = quantile(`Category Norm`, 0.75, na.rm = TRUE),
    IQR = IQR(`Category Norm`, na.rm = TRUE),
    lower = Q1 - 1.5 * IQR,
    upper = Q3 + 1.5 * IQR,
    is_outlier = `Category Norm` < lower | `Category Norm` > upper
  ) %>%
  ungroup()

#plot
item_wise_ratings_plot<-stimuli %>%
  filter(!is.na(Category), Category != "NA") %>%
  ggplot(aes(x = Category, y = `Category Norm`, fill = Category)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
  coord_flip() + scale_y_reverse()+
  geom_text_repel(
    data = stimuli_outliers %>% filter(is_outlier),
    aes(label = Name),
    size = 3,
    max.overlaps = 50
  ) +
  theme_classic() + ylab("Category Fit (Poorest Fit -> Best Fit)")+ggtitle("Distribution of Category Norm Ratings")+
  theme(legend.position = "none")

item_wise_ratings_plot

#Item wise distribution of label x item similarity across categories (DSM derived)
#ID outliers
interitem_outliers <- item_similarities_long %>%
  filter(Category != "NA") %>%
  group_by(Category) %>%
  mutate(
    Q1 = quantile(`Similarity`, 0.25, na.rm = TRUE),
    Q3 = quantile(`Similarity`, 0.75, na.rm = TRUE),
    IQR = IQR(`Similarity`, na.rm = TRUE),
    lower = Q1 - 1.5 * IQR,
    upper = Q3 + 1.5 * IQR,
    is_outlier = `Similarity` < lower | `Similarity` > upper
  ) %>%
  ungroup()

#plot
itemwise_dsm_plot<-item_similarities_long %>%
  filter(!is.na(Category), Category != "NA") %>%
  ggplot(aes(x = Category, y = `Similarity`, fill = Category)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
  coord_flip() +
  geom_text_repel(
    data = interitem_outliers %>% filter(is_outlier),
    aes(label = Item),
    size = 3,
    max.overlaps = 50
  ) +
  theme_classic() +ylab("Category Fit (Poorest Fit -> Best Fit)")+ggtitle("Distribution of Label x Item Similarities")+
  theme(legend.position = "none")

itemwise_dsm_plot
```

But we can also look at the distribution of item-item similarities and identify items with weak cohesion. Below, I've marked (in red) the 3 items in each category with the lowest average similarity.

```{r}

base_dir <- "stimuli_by_category"
cat_dirs <- list.dirs(base_dir, recursive = FALSE)

walk(cat_dirs, function(cat_path) {
  cat_name <- basename(cat_path)
  file_path <- file.path(cat_path, "USE_semantic_matrix_items.csv")

  # read matrix
  mat <- read.csv(file_path, row.names = 1, check.names = FALSE)
  mat <- as.matrix(mat)
  mode(mat) <- "numeric"
  
  # compute mean similarity (exclude diagonal)
  mat_avg <- mat
  diag(mat_avg) <- NA
  item_means <- rowMeans(mat_avg, na.rm = TRUE)

  # identify 3 lowest items
  lowest_items <- names(sort(item_means))[1:3]
  label_colors <- ifelse(rownames(mat) %in% lowest_items, "red", "black")
  
  col_fun <- colorRampPalette(c("red", "navy"))(100)
  
  # plot
  corrplot(
    mat,
    method = "color",
    type = "full",
    col = col_fun,
    tl.col = label_colors,
    tl.cex = 0.7,
    title = cat_name,
    col.lim = c(0, 1),
    mar = c(0, 0, 2, 0))})

```

